---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# msnz

The goal of the `msnz` package is to provide functions to simplify data 
analysis from the New Zealand Multiple Sclerosis Prevalence and related studies.

## Installation

This package is currently only of interest and utility at NZBRI and hence won't 
 be made available via CRAN. Instead, it is hosted in a repository on 
 Github at https://github.com/nzbri/msnz
 
 Therefore, the usual installation route using `install.packages('msnz')` is 
 not possible, and will yield a not-useful error message that the package is not
 available for your version of R. Instead, install `msnz` from its development
 repository on Github as follows:

```{r gh-installation, eval = FALSE}
# install.packages('remotes')
remotes::install_github('nzbri/msnz')
```
If `install_github('nzbri/msnz')` is invoked subsequently, the package will be
downloaded and installed only if the version on Github is newer than the one 
installed locally.

If problems arise in a new release, you can downgrade to a previous version by 
specifying the name of a particular release to revert to, e.g.

```{r gh-revert, eval = FALSE}
remotes::install_github('nzbri/chchpd@v0.2.0')
```
