---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# msnz

The goal of the `msnz` package is to provide functions to simplify data 
analysis from the New Zealand Multiple Sclerosis Prevalence and related studies,
for researchers associated with the New Zealand Brain Research Institute 
(NZBRI).

The package also provides a function to estimate life expectancy from the New 
Zealand cohort life tables, which can be of use more widely. These individual 
estimates are either: 

- Deterministic - the function simply looks up the expected year of death for
an individual of a given sex, year of birth, and conditional age, or
- Probabilistic - the function runs an individual-level random simulation for a
person of given characteristics. This means that the generated sample will have
a realistic level of variability across individuals, rather than people with the
same demographics having an identical expected year of death.

## Installation

This package is not available via CRAN and is hosted in a repository on Github 
at https://github.com/nzbri/msnz
 
Therefore, the usual installation route using `install.packages('msnz')` is 
not possible, and will yield a not-useful error message that the package is not
available for your version of R. Instead, install `msnz` from its development
repository on Github as follows:

```{r gh-installation, eval = FALSE}
# install.packages('remotes')
remotes::install_github('nzbri/msnz')
```
If `install_github('nzbri/msnz')` is invoked subsequently, the package will be
downloaded and installed only if the version on Github is newer than the one 
installed locally.

If problems arise in a new release, you can downgrade to a previous version by 
specifying the name of a particular release to revert to, e.g.

```{r gh-revert, eval = FALSE}
remotes::install_github('nzbri/msnz@v0.3.0')
```

## Functions specific to the MS prevalence study

The package provides three convenience functions to extract info from the unique
identifier (UIN) assigned to each participant (see Richardson _et.al._ (2012)
Method for identifying eligible individuals for a prevalence survey in the 
absence of a disease register or population register. 
_Internal Medicine Journal,_ ***42,*** 1207-1212):

```{r uni-functions, eval = TRUE}
library(msnz)

msnz::dob_from_uin('01011970FAB')

msnz::sex_from_uin('01011970FAB')

msnz::initials_from_uin('01011970FAB')
```

It also provides two package-wide constants, giving the original census date of 
the New Zealand MS Prevalence study, and the censoring date for survival 
analyses, set to 15 years later:

```{r date-constants, eval = TRUE}
msnz::census_date

msnz::censoring_date
```

## Calculating New Zealand life expectancy

This function is the only one of use more generally to external researchers. The
package incorporates the New Zealand Cohort Life Tables provided by 
Statistics New Zealand, as released in March 2021. (See https://www.stats.govt.nz/information-releases/new-zealand-cohort-life-tables-march-2022-update
for source data).

The `msnz::expected_year_of_death()` function allows one to calculate the life 
expectancy of a New Zealander, given their year of birth, sex, and some 
conditional age (see below for explanation).

```{r example-1, eval = TRUE}
msnz::expected_year_of_death(year_of_birth = 1970, 
                             sex = 'female', 
                             conditional_age = 50)

# calculate life expectancy (in years from the conditional age year):
year_of_birth = 1970
conditional_age = 50

expected_year_of_death(year_of_birth = year_of_birth,
                       sex = 'female', 
                       conditional_age = conditional_age) - 
  (year_of_birth + conditional_age) # = 2020

```
It is important to specify a conditional age that the person has reached. If 
`0`, then the returned value will reflect life expectancy at birth. For example,
the expected year of death of a person is greater if it is known that they have 
managed to reach a given age, compared to their life expectancy at birth (when
they would be subject to child mortality and early adulthood risks):

```{r example-2, eval = TRUE}
# life expectancy at birth, which is subject to infant mortality and elevated 
# risks in teenage years/early adulthood (e.g. traffic accidents and suicide):
msnz::expected_year_of_death(year_of_birth = 1970, 'female', 
                             conditional_age = 0)

# given that we know that a person has lived to a certain age, their expected 
# year of death should be greater, as they have survived through a period of
# mortality risks:
msnz::expected_year_of_death(year_of_birth = 1970, 'female', 
                             conditional_age = 50)

```

The values returned can either be deterministic (extracted directly from the 
life table) or the result of a random simulation process (to better approximate 
the distribution of values that would be seen in an actual population). The 
simulation approach is particularly useful in a survival analysis where one 
wishes to compare the survival in a given sample against a synthetic sample 
derived from the population. That is, for each person in the actual sample, we
generate a synthetic comparison person, randomly simulated from the population, 
conditional on the target person's year of birth, sex, and age at a censoring 
date. The simulation approach gives a much more natural-looking population 
comparison survival distribution, compared to each synthetic person having 
precisely the mean life expectancy conditional on those values.

The simulation process is invoked by specifying the parameter 
`method = 'sample'`, rather than the default value `method = 'median'`, which 
simply returns the tabulated median value from the cohort life tables. Using the
`'sample'` method runs a simulation for a person of the given year of birth, 
sex, and conditional age (e.g. the age they reached at the census date). The
simulation proceeds by iterating up from that conditional age, one year at a
time. At each year of age, a random number is generated (uniformly distributed 
between 0.0 and 1.0). If that number is less than the life table probability of 
such a person surviving to the next year, the age is incremented and the 
simulation continues for that individual. If the random number exceeds that 
probability, then the current age is assigned as the synthetic person’s age of 
death.

For fuller details on other function parameters, look up the help via 
`?expected_year_of_death`. These include control over what estimate is returned. 
(i.e. the median or the 5th or 95th percentile), the number of simulated samples
returned per individual, and a seed to allow for reproducibility of the randomly
simulated values.
